Longitudinal Principal Manifold Estimation

### 2023-10-11 paper/enar_2024/enar_submission_2024.tex
add abstract

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}

%\usepackage{algorithm2e}

@@

\newcommand{\kmedit}[1]{{\color{purple}  #1}}
\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}

\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}

@@

\title[This is an Example of Recto Running Head]{This is an Example of
an Article Title}

@@

\author
{John Author\emailx{author@address.edu} \\
Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K. 
\and
Kathy Author\emailx{anotherauthor@address.edu} \\
Department of Biostatistics, University of North Carolina at Chapel Hill, 
Chapel Hill, North Carolina, U.S.A.}

@@

\date{{\it Received October} 2007. {\it Revised February} 2008.  {\it
Accepted March} 2008.}

@@

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{64}
\pubyear{2008}
\artmonth{December}

@@

\doi{10.1111/j.1541-0420.2005.00454.x}

@@

\section{Introduction}
\label{s:intro}

Here is the introduction. The file \texttt{biomsample.tex} gives an
example of using the \texttt{natbib} package to cross-reference
citations from the bibliography.  Here, we just do this manually to
demonstrate that the old-fashioned way also is acceptable.  See the
comment right before the bibliography section below for information on
using BiBTeX.

Please note that, although this document class produces a final
product that is very close to the format and appearance of a typeset
\textit{Biometrics} article, there may be some idiosyncracies that
cause the format to deviate slightly from that in the journal.  These
will be corrected at the typesetting phase should your paper be accepted
and forwarded for publication.  So do not worry if such things occur!

\section{Model}
\label{s:model}

\subsection{First Model Subsection}

The Cox model (Cox, 1972) is one of the most widely used statistical
models.  Hastie, Tibshirani, and Friedman (2001) is an example of a
citation to a work with three authors.  The first time you reference
one of these in the text, use all the authors names. However, in all
subsequent references, just use Hastie et al. (2001).  Works with four
or more authors are always referenced in the text using ``et al.''
All authors names should appear in the bibliography for all entries.

\subsection{Second Model Subsection}

Please use a recent issue of \textit{Biometrics} as a guide to the style
for citations and bibliography entries, and follow that style exactly!!

\section{Inference}
\label{s:inf}

Please see the file \texttt{biomsample.tex} for fancy examples of making
tables.  Here is a very simple one.  Use \texttt{table} for tables
that are narrow enough to fit in one column of the typeset journl; use
\texttt{table*} for tables that need to span two columns.  For
figures, use of \texttt{figure} and \texttt{figure*} is analogous.

\begin{table}
\caption{This is a simple table.}
\label{t:one}
\begin{center}
\begin{tabular}{lrrr}
\Hline
Estimator & \multicolumn{1}{c}{$\beta_1$} &  \multicolumn{1}{c}{$\beta_2$} & 
\multicolumn{1}{c}{$\beta_3$} \\ \hline
MLE & 10.18 & $-$3.26 & 0.13 \\
OLS & 9.92 & $-$3.19 & 0.11 \\
WLS & 9.88 & $-$3.33 & 0.12 \\
\hline
\end{tabular}
\end{center}
\end{table}

You can experiment with fancier tables than Table~\ref{t:one}.

We can get bold symbols using \verb+\bmath+, for example, $\bmath{\alpha}_i$.

\section{Discussion}
\label{s:discuss}

Put your final comments here. Here I test what happens when I cite \cite{mengPrincipalManifoldEstimation2021}.

%  The \backmatter command formats the subsequent headings so that they
%  are in the journal style.  Please keep this command in your document
%  in this position, right after the final section of the main part of 
%  the paper and right before the Acknowledgements, Supporting Information (Supplementary %  Materials),   and References sections.

>>>>

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
% \usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}

\usepackage{algorithm2e}

@@

@@

\title[Longitudinal Principal Manifold Estimation]{Longitudinal Principal Manifold Estimation}

@@

@@

@@

@@

@@

\section{Introduction}

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}). Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}). For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes. Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}). Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}). While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained. In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.

Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time. Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum. To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits. Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus in individuals with Alzheimer's disease (AD). The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

\section{Principal Manifolds and Estimation Algorithm}\label{s:PME}

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings requires is problematic. Motivated by the penalization approaches presented by in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, i.e. $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows. 
\begin{definition}
  \label{def:principal_manifolds}
  Given a random $D$-vector $\boldsymbol{X}$ associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments. Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty]$, and the following functionals:
  \begin{align}\label{eq:5}
&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
  \end{align}
  The principal manifold of $\boldsymbol{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if 
  \begin{equation}\nonumber
    \begin{aligned}
        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f),
    \end{aligned}
  \end{equation}
where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$. 
\end{definition}
The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements. First is the expected distance from each point in the cloud to the fitted manifold. In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on 3-dimensional brain region of interest. While the second element of the equation imposes the smoothness penalty on the estimated manifold. The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$. In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. It is important to note, that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.

\section{Longitudinal Principal Manifold Estimation Algorithm}\label{s:LPME}

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally, during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e. modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$ which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of the existing approaches for the setting described above, we now propose longitudinal principal manifold estimation (LPME) algorithm to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

To give a more general representation of this problem, let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. However, in this updated case, the function determining the manifold may change over time.

\subsection{The Longitudinal Principal Manifold Framework}

Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively. For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g. the surface of the 3D brain region, at time $t$. 
\begin{definition}
  \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows. 
\begin{align}\label{eq:newKappa}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt. \nonumber
\end{align}
where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}
By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, while imposing smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$. These goals correspond to minizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively. Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions. Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together. Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa. Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered to be smooth over the other set of variables.

\subsection{The LPME Algorithm}

The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.

\subsection*{Data Reduction}
We begin by considering $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} (Section 5.3 therein) to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$. HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.

This step reduces the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to the size $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

\subsection*{Initialization}
The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.

Instead of using Isomap on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one Isomap run to reach a parameterization for the centers across all time points simultaneously. Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.

\subsection*{Fitting}
Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$, which deals with the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in Eq.~\eqref{eq:11}.

The remainder of the fitting stage deals with the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$, i.e., the curvature in the time dimension (see Eq.~\eqref{eq:11}). Specifically, the remainder of this stage involves estimating a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.

At each time point, the estimated principal manifold has a spline form. As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points being used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ being specified based on the maximum number of clusters identified in the data reduction step at any time point. This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
\begin{itemize}
    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
\end{itemize}
This minimization is conducted by solving
\begin{equation}
  \left(
    \begin{array}{cc}
      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
      \boldsymbol{R} & \boldsymbol{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      \boldsymbol{s}_t \\
      \boldsymbol{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \boldsymbol{Y}_t \\
      \boldsymbol{0}
    \end{array}
  \right). \label{eq:13}
\end{equation}
The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
\begin{equation}
  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
\end{equation}

We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$. To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to 
$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$

With a given tuning value $\gamma$, we the fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$. This function takes the form

\begin{equation}
  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
\end{equation}
where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to 
\begin{equation}
  \left(
  \begin{array}{ccc}
    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{A} + 2\gamma\boldsymbol{A} & 2\boldsymbol{A}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{T} \\
    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{A} & 2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{0} \\
    \boldsymbol{T}^{T} & \boldsymbol{0} & \boldsymbol{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \boldsymbol{\delta} \\
    \boldsymbol{\nu} \\
    \boldsymbol{m}
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{\Omega} \\
    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{\Omega} \\
    \boldsymbol{0}
  \end{array}
  \right), \label{eq:16}
\end{equation}
where $\boldsymbol{\Omega}=(\boldsymbol{\omega}_1,\ldots, \boldsymbol{\omega}_T)^\T$.

With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$. Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\boldsymbol{r}$ may be written as
\begin{equation}
  F_{\gamma}(t, \boldsymbol{r}) = \sum_{j=1}^{N}\boldsymbol{s}_{\gamma}(t)_j \eta_{d}\left(\|\boldsymbol{r} - \boldsymbol{r}_j^*\|\right) + \sum_{k=1}^{d+1}\boldsymbol{\alpha}_{\gamma}(t)_k \phi_k(\boldsymbol{r}). \label{eq:17}
\end{equation}

\subsection*{Tuning}

Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows: 
\begin{equation}
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
\end{equation}
The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.

Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.

\RestyleAlgo{ruled}
\LinesNumbered
